Palomar 5 Gaps

### 2024-09-23 paper/main.tex
editted captions and such

  Stellar streams, long and thin structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster’s orbit. Theoretical predictions of this phenomenon existed long before the first observation in 1995 by \citet{carl_j_grillmair_globular_1995}, who detected stars beyond the tidal radii of globular clusters. Subsequent work by \citet{odenkirchen_detection_2001} further demonstrated that Palomar 5’s stars not only extended beyond the tidal radius but also formed distinct tidal tails.

  Over the past two decades, theoretical and numerical studies have shown that stellar streams can constrain the gravitational field of host galaxies, particularly the Milky Way. By applying Poisson’s equation, the density distribution and gravitational potential can be separated into different components, including the dark matter halo. This halo, often modeled as spherical, is the largest component and crucial to explaining the constant rotational velocity of galaxies. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. Meanwhile, Bonaca et al. employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we briefly summarize here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by Baumgardt and Vasiliev et al. We handle the uncertainties through a monte-carlo approach. The catalog provides uncertainties for distances, radial velocities, proper motions, and mass, and a covariance term between the proper motions. We repeate the whole experiment 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. However, for the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

@@

    For the galactic potential, we employed the second static model proposed by Pouliasis et al. (2017), which consists of a superposition of a thin disk, thick disk, and a dark matter halo. The values are provided in Table 1 of Ferrone et al 2023. This model is time-independent over the course of our simulations.

@@

    We then integrated the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as in eq.~\ref{eq:GCNBody}; instead, at each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right).
      \end{equation}

@@

    During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved 5,000 intermediate timesteps which is a temporal resolution of 1 million years. To provide an estimate of the size of our simulations, we have $N_p=100 000$, $N_\textrm{time stamps} = 5000$, $N_\textrm{phase}=6$, $N_\textrm{sampling}=50$, and using the data type as a single precision floating point number $\textrm{Single}=4$~bytes results in approximately 600 GB of data.

@@

      This method, however, has its limitations, especially when detecting smaller gaps. As outlined by Erkal et al. (2015), a gap's size may not indicate a weak impact, but rather a more recent one. Additionally, since our streams have finite width, some gaps are oblique with respect to the stream axis. In such cases, marginalizing over $y'$ erases the gap's signal, making it impossible to detect in a 1D profile. This limitation is particularly evident in gaps caused by NGC2808, as discussed in the results. Therefore, this quantitative analysis serves as an \textit{aid} to visual inspection rather than a complete substitute for it. This method helps particularly with large subtle gaps that the eye does not notice in the 2D maps.

@@

      A key question we seek to answer is: from the gaps present at the end of the simulation, who caused them? To address this, we examine the evolution of stream density over time. Instead of using the x'coordinate, we introduce $\tau$, which represents time rather than distance. Specifically, $\tau$ indicates how long it will take for a cluster to reach, or how long ago it passed, a given point on its orbit. This is advantageous because the growth of the stream is approximately linear in $\tau$ whereas in physical space, streams on eccentric orbits expand and contract depending on the orbital phase.

@@

      We then used \texttt{scipy}'s \texttt{ndimage} package to identify the top five local maxima in the data space of gravitational acceleration $\vec{g}$ as a function of time $t$ and the stream coordinate $\tau$. Once these local maxima were identified, we iterated over the contributions of individual globular clusters to determine which cluster contributed the most to each peak in $\vec{g}$. Each significant peak was labeled with the corresponding globular cluster.

      Afterward, we cross-referenced these peaks with the previously identified gaps in the stream. For large gaps resulting from strong interactions, we observed that after an impact, a low-density wake is left behind in the ($t,\tau$) plane.

@@

  \subsection{Reconstruction of the impact geometry}
    In its simplest form, impact theory, as presented in works such as Binney and Tremaine, John Taylor, and Galaxies, considers two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is often employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption greatly simplifies the computation.

@@

    To achieve this, we begin by identifying the approximate moments of impact from the most significant clusters, as determined in the previous analysis in Sec.~\ref{SuspectToCulprit}. Then, we refine these estimates. While the orbit serves as a good first-order proxy for the stream, it is insufficient, as the typical offset between the stream and its orbit can be on the order of 200-400 pc—significantly larger than the impact parameters we are investigating.

    In this refinement process, we aim to pinpoint both the exact location of the impact along the stream and the precise moment it occurred. To do so, we fit a third-order parametric polynomial to the stream, using the saved snapshots from our simulations:
    \begin{equation}
      \vec{s}(\tau) = 
      \left\{
        \begin{aligned}
          x(\tau) &= a_0 + a_1 \tau + a_2 \tau^2 + a_3 \tau^3 \\ 
          y(\tau) &= b_0 + b_1 \tau + b_2 \tau^2 + b_3 \tau^3 \\
          z(\tau) &= c_0 + c_1 \tau + c_2 \tau^2 + c_3 \tau^3
        \end{aligned}
      \right.
      \end{equation}  
    where $x$, $y$, and $z$ represent the parametric line describing the stream in galactocentric coordinates, $\tau$ is the stream coordinate in time as described in the previous section, and is used as the independent variable in this analysis. The coefficients $a_i$, $b_i$, and $c_i$ are the polynomial coefficients. We found that a second-order polynomial was insufficient to capture the curvature along the full length of the stream, with divergence from the star particle positions occurring at the extremes. A third-order polynomial was sufficient and desirable, as it is the lowest order that adequately captures the path of the stream over the entire length under consideration.

    In this analysis, only one side of the stream is considered. For instance, if the impact candidate was in the leading tail, only the star particles with $\tau > 0$ are used to constrain the stream. The polynomial coefficients were determined using the Nelder-Mead algorithm from \texttt{scipy}'s optimization package.

    Since the simulation snapshots were saved at a temporal resolution of 1 Myr—rather than at the integration timestep of $10^4$ years, which would generate an excessive amount of data—we interpolate between snapshots to more precisely estimate the impact geometry. This is achieved by fitting the stream at five timesteps surrounding the approximate impact time, a period of 5 Myr, which sufficiently covers the interaction time. The interaction time can be estimated as $t \approx \frac{100~\text{pc}}{300~\frac{\text{km}}{\text{s}}} \approx 0.3~\text{Myr}$.

@@

  \begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{force_on_orbit-monte-carlo-009.png}
    \caption[]{This figure demonstrates how we determined which globular clusters were responsible for the gaps. The bottom plot showcases the evolution of the stream density in simulation time. The y-axis is $\tau$, which is a coordinate in units of time that indicates how far ahead or behind it is from a globular cluster. The x-axis is the simulation time, $t$, where 0~s km~kpc$^{-1}$ indicates present time. The density was used to determine a sutible length of the stream as a function of time. This length was then used to extract a piece of Palomar 5's orbit about a given position of Pal 5 in time. This orbital segment is used to approximate the stream. Then, the gravitational force from all other clusers was computed on the orbit. Moments of high acceleration indicate the passage of another cluster. The top 5 strongest passages are labeled with red circles as well as the name of the clusters. }
    \label{fig:force-on-orbit}
    \end{figure*}

  \begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps.png}
    \caption[]{This figure decomposes the gaps found in the full simulation to the responsible perturbers. The pertuber candidates were found in Fig.~\ref{fig:profiles}, and extra simulations were run in which we singularly add the suspects to verify that it was indeed them who created the gaps. The titles explain which simulation is which, and there's also a descrepandy map. (I think I'm going to get rid of the DIFF map)}
    \label{fig:decomposition}
    \end{figure*}

>>>>

  Stellar streams, long and thin structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit. Theoretical predictions of this phenomenon existed long before the first observation in 1995 by \citet{carl_j_grillmair_globular_1995}, who detected stars beyond the tidal radii of globular clusters. Subsequent work by \citet{odenkirchen_detection_2001} further demonstrated that Palomar 5's stars not only extended beyond the tidal radius but also formed distinct tidal tails.

  Over the past two decades, theoretical and numerical studies have shown that stellar streams can constrain the gravitational field of host galaxies, particularly the Milky Way. By applying Poisson's equation, the density distribution and gravitational potential can be separated into different components, including the dark matter halo. This halo, often modeled as spherical, is the largest component and crucial to explaining the constant rotational velocity of galaxies. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. Meanwhile, Bonaca et al. employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by Baumgardt and Vasiliev et al. We handle the uncertainties through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions, with a covariance term between the proper motions. We repeate the whole experiment 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. However, for the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

@@

    For the galactic potential, we employed the second static model proposed by Pouliasis et al. (2017), which consists of a superposition of a thin disk, thick disk, and a dark matter halo. The masses and scale lengths are provided in Table 1 of Ferrone et al 2023. This model is time-independent over the course of our simulations.

@@

    We then integrated the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as in eq.~\ref{eq:GCNBody}; instead, at each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right).
      \end{equation} \label{eq:equation_of_motion_particle}

@@

    During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved $5000$ intermediate timesteps which is a temporal resolution of 1 million years. To provide an estimate of the size of our simulations, we have $N_p=100 000$, $N_\textrm{time stamps} = 5 000$, $N_\textrm{phase}=6$, $N_\textrm{sampling}=50$, and using single precision floating point numbers, $\textrm{Single}=4$~bytes, results in approximately 600 GB of data.

@@

      This method, however, has its limitations, especially when detecting smaller gaps. As outlined by Erkal et al. (2015), a small gap is indicative of a weak impact, but rather a more recent one. This is because gap growth is a dispersion phenomenon. Additionally, since our streams have finite width, some gaps are oblique with respect to the stream axis. In such cases, marginalizing over $y'$ erases the gap's signal, making it impossible to detect in a 1D profile. This limitation is particularly evident in gaps caused by NGC~2808, as discussed in the results. Therefore, this quantitative analysis serves as an \textit{aid} to visual inspection rather than a complete substitute for it. This method helps particularly with large subtle gaps that the eye does not notice in the 2D maps.

@@

      A key question we seek to answer is: from the gaps present at the end of the simulation, who caused them? To address this, we examine the evolution of stream density over time. Instead of using the x'-coordinate, we introduce $\tau$, which represents time rather than distance. Specifically, $\tau$ indicates how long it will take for a cluster to reach, or how long ago it passed, a given point on its orbit. This is advantageous because the growth of the stream is approximately linear in $\tau$ whereas in physical space, streams on eccentric orbits expand and contract depending on the orbital phase.

@@

      \begin{figure*}
        \centering
        \includegraphics[width=\linewidth]{force_on_orbit-monte-carlo-009.png}
        \caption{This figure demonstrates how we determined which globular clusters were responsible for the gaps. The y-axis is $\tau$, which is a coordinate in units of time that indicates how far ahead or behind it is from a globular cluster. The x-axis is the simulation time, $t$, where 0~s km~kpc$^{-1}$ indicates present time.  The bottom plot showcases the evolution of the stream density in simulation time. The density was used to determine a sutible length of the stream as a function of time. This length was then used to extract a piece of Palomar 5's orbit about a given position of Pal 5 in time. This orbital segment is used to approximate the stream. Then, the gravitational force from all other clusers was computed on the orbit. This is shown in the top plot, where the gravitational force is in measured in acceleration and is given in integration units: km$^2$~kpc$^{-1}$~s$^{-2}$. Moments of high acceleration indicate the passage of another cluster. The top 5 strongest passages are labeled with red circles as well as the name of the clusters. The example shown in this plot is the same simulation as Fig.~\ref{fig:stream_on_sky}.}
        \label{fig:force-on-orbit}
        \end{figure*}

      We then used \texttt{scipy}'s \texttt{ndimage} package to identify the top five local maxima in the data space of gravitational acceleration $\vec{g}$ as a function of time $t$ and the stream coordinate $\tau$. TheThis is done by first smoothing the image by taking a 5 point moving average kernel. Secondly, we use maximum filter to locate coordinates in the $t,\tau$ data plane who are local maxima to at least 10-adjacent data points. These locations are then ordered and the top five strongest interactions are saved. Once these local maxima were identified, we iterated over the contributions of individual globular clusters to determine which cluster contributed the most to each peak in $\vec{g}$. Each significant peak was labeled with the corresponding globular cluster.

      Afterward, we cross-referenced these peaks with the locations of the gaps identified by studying the density maps and profiles from Fig.~\ref{fig:profiles}. For large gaps resulting from strong interactions, we observed that after an impact, a low-density wake is left behind in the ($t,\tau$) plane, which can be seen corresponding to the impacts of NGC~104 and NGC~7808 in the bottom panel of Fig.~\ref{fig:force-on-orbit}.

@@

      \begin{figure*}
        \centering
        \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps.png}
        \caption{This figure decomposes the gaps shown in the middle panel of Fig.~\ref{fig:profiles} into the individual perturbuers whose cause them, as identified from Fig.~\ref{fig:force-on-orbit}. These simulations only include the suspected globular clusters in one-at-a-time.}
        \label{fig:decomposition}
        \end{figure*}

  \subsection{Reconstruction of the impact geometry}

    With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster and a stream to have in order for a gap to form. We turn to impact theory, which in its simplest form as presented in works such as Binney and Tremaine, John Taylor, and Galaxiesbook, considers two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is often employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption greatly simplifies the computation.

@@

    To achieve this, we begin by identifying the approximate moments of impact from the most significant clusters, as determined in the previous analysis in Sec.~\ref{SuspectToCulprit}. Then, we refine these estimates. While the orbit serves as a good first-order proxy for the stream, it is insufficient, as the typical offset between the stream and its orbit can be on the order of 200-400 pc—--significantly larger than the impact parameters we are investigating.

    In this refinement process, we aim to pinpoint both the exact location of the impact along the stream and the precise moment it occurred. To do so, we fit a third-order parametric polynomial to the stream, using the saved snapshots from our simulations:
    \begin{equation}
      \vec{s}(\tau) = 
      \left\{
        \begin{aligned}
          x(\tau) &= a_0 + a_1 \tau + a_2 \tau^2 + a_3 \tau^3 \\ 
          y(\tau) &= b_0 + b_1 \tau + b_2 \tau^2 + b_3 \tau^3 \\
          z(\tau) &= c_0 + c_1 \tau + c_2 \tau^2 + c_3 \tau^3
        \end{aligned}
      \right.
      \end{equation}  
    where $x$, $y$, and $z$ represent the parametric line describing the stream in galactocentric coordinates, $\tau$ is the stream coordinate in time as described in the previous section, and is used as the independent variable to parameterize the motion. The coefficients $a_i$, $b_i$, and $c_i$ are the polynomial coefficients. We found that a second-order polynomial was insufficient to capture the curvature along the full length of the stream, with divergence from the star particle positions occurring at the extremes. A third-order polynomial was sufficient and desirable, as it is the lowest order that adequately captures the path of the stream over the entire length under consideration.

    In this analysis, only one side of the stream is considered. For instance, if the impact candidate was in the leading tail, only the star particles with $\tau > 0$ are used to constrain the stream track. The polynomial coefficients were determined using the through a minimization method using Nelder-Mead algorithm from \texttt{scipy}'s optimization package.

    Since the simulation snapshots were saved at a temporal resolution of 1 Myr--—rather than at the integration timestep, which would have generated an excessive amount of data—--we interpolate between snapshots to more precisely estimate the impact geometry. This is achieved by fitting the stream at five timesteps surrounding the approximate impact time, a period of 5 Myr, which sufficiently covers the interaction time. The interaction time can be estimated as $t \approx \frac{100~\text{pc}}{300~\frac{\text{km}}{\text{s}}} \approx 0.3~\text{Myr}$.

@@
